---
title: "QuantDev1"
output: html_notebook
---
From
https://quantdev.ssri.psu.edu/tutorials/intro-data-reshape-plots-wisc

```{r packages}
library(psych)    #for general use functions
library(ggplot2)  #for plotting
library(car)      #for general use functions
library(GGally)   #for a specicif plot
library(lattice)  #for plotting
```
```{r readdat}
############################
####### Reading in the Data
############################
#set filepath for data file
#this has worked but seems dodgy now so read direct from file
#but command syntax should be useful so retain
readfromweb<-0
if(readfromweb==1){
filepath <- "https://quantdev.ssri.psu.edu/sites/qdev/files/wisc3raw.csv"
#read in the .csv file using the url() function
wisc3raw <- read.csv(file=url(filepath),header=TRUE)
}
if(readfromweb==0)
{
 wisc3raw<- read.csv('wisc3raw.csv')
}


```

Data cleaning etc  
Some useful tips/tricks here

```{r dataclean}
colnames(wisc3raw)
dput(colnames(wisc3raw))  #put into a format that is easy to cut and paste

# Set all variable names to lowercase 
var_names <- tolower(colnames(wisc3raw))
colnames(wisc3raw)<-var_names


#Output a comma delimited data file;
write.csv(wisc3raw, file="NewData.csv",row.names=FALSE, na="")
```


## Manipulating the Repeated Measures Data (Long Data and Wide Data)  

(I'm skipping things I am v familiar with here).

```{r subset}


#First, lets subset down to the variables we need.

#making a vector of variable names (id, time-varying, time-invariant)
var_names_sub <- c("id", 
                   "verb1","verb2","verb4","verb6",
                   "perfo1","perfo2","perfo4","perfo6",
                   "momed","grad")
#subsetting
wiscraw <- wisc3raw[,var_names_sub]

```

```{r reshaping}
#reshaping wide to long
wisclong <- reshape(data=wiscraw,
                    varying = c("verb1","verb2","verb4","verb6",
                                "perfo1","perfo2","perfo4","perfo6"),
                    timevar=c("grade"), 
                    idvar=c("id"),
                    direction="long", sep="")
#sorting for easy viewing
#reorder by id and day
wisclong <- wisclong[order(wisclong$id,wisclong$grade), ]

#looking at the data
head(wisclong, 8)

```

reverse - reshape long to wide
```{r reshapeagain}
#reshaping long to wide
wiscwide <- reshape(data=wisclong, 
                    timevar=c("grade"), 
                    idvar=c("id"),
                    v.names=c("verb","perfo"),
                    direction="wide", sep="_")

#reordering columns for easy viewing
wiscwide <- wiscwide[,c("id",
                        "verb_1","verb_2","verb_4","verb_6",
                        "perfo_1","perfo_2","perfo_4","perfo_6",
                        "momed","grad")]

#looking at the data
head(wiscwide)
```

A more general reshaping solution is provided by Hadley Wickham’s reshape2 package through melt and cast functions.

```{r basicdescriptives}
describe(wiscwide)
```

for repeated measures...

```{r descrepeated}
describe(wisclong$verb)
#here ignoring time
#histogram
ggplot(data=wisclong, aes(x=verb, y=..density..)) +
  geom_histogram(binwidth=2.5, fill="white", color="black") + 
  geom_density(color="red") +
  xlab("Verbal Ability (Grade 1 to 6)")
#sample descriptives by occasion 
#in the wide file
describe(wiscwide[,c("verb_1","verb_2","verb_4","verb_6")])
#histogram faceted by grade
ggplot(data=wisclong, aes(x=verb)) +
  geom_histogram(binwidth=5, pad = TRUE, fill="white", color="black") + 
  xlab("Verbal Ability") +
  facet_grid(grade ~ .)

#boxplot by grade
qplot(x=factor(grade), y=verb, data=wisclong, geom="boxplot", ylab="Verbal Ability", xlab="Grade")

#use factor() to convert "time" from numeric to categorical

#boxplot by grade
ggplot(data=wisclong, aes(x=factor(grade), y=verb)) + 
  geom_boxplot(notch = TRUE) +
  stat_summary(fun.y="mean", geom="point", shape=23, size=3, fill="white") +
  labs(x = "Grade", y = "Verbal Ability")

#Density distribution by grade
ggplot(data=wisclong, aes(x=verb)) + 
  geom_density(aes(group=factor(grade), colour=factor(grade), fill=factor(grade)), alpha=0.3) +
  guides(colour=FALSE,
         fill=guide_legend(title="Grade")) +
  labs(x="Verbal Ability", y="Density")
```
Above, we looked at the means and variances. Because these are repeated measures, we also have covariances.

```{r cors}
# Correlations 
cor(wiscwide[,c("verb_1","verb_2","verb_4","verb_6")], use="complete.obs",method="spearman")
#Correlations plot
pairs(wiscwide[,c("verb_1","verb_2","verb_4","verb_6")])

#in the psych library
pairs.panels(wiscwide[,c("verb_1","verb_2","verb_4","verb_6")])


```
## 3.2.3 Individual-level descriptives across Time (Verbal Ability)  

Note that our interest is often in individual development, rather than sample development. We need to consider how each individual is changing over time.

Thus, we are interested in Verbal Ability across Time = individual persons.  

```{r sect3.2.3}

#Using library(lattice)
#Plotting intraindividual change
xyplot(verb ~ grade, groups=id, 
       data=wisclong, type="l",
       main="Verbal Ability Trajectories")

#Using library(ggplot2) ... see also http://ggplot.yhathq.com/docs/index.html
#Plotting intraindividual change
ggplot(data = wisclong, aes(x = grade, y = verb, group = id)) +
  geom_point() + 
  geom_line() +
  xlab("Grade") + 
  ylab("Verbal Ability") + ylim(0,80) +
  scale_x_continuous(breaks=seq(1,6,by=1))

#plotting just a subset with colour
ggplot(data = wisclong[which(wisclong$id <=20),], aes(x = grade, y = verb, group = id, color=factor(id))) +
  geom_point() + 
  geom_line(data=wisclong[which(wisclong$id <= 20 & wisclong$verb !="NA"),]) +
  xlab("Grade") + 
  ylab("Verbal Ability") + ylim(0,80) +
  scale_x_continuous(breaks=seq(1,6,by=1)) +
  guides(color=FALSE)

#old-school style using library(lattice)
xyplot(verb~grade|id, data=wisclong[wisclong$id<21,], as.table=TRUE)

#ggplot2 style
ggplot(data = wisclong[which(wisclong$id <=20),], aes(x = grade, y = verb, group = id)) +
  geom_point() + 
  geom_line(data=wisclong[which(wisclong$id <= 20 & wisclong$verb !="NA"),]) +
  xlab("Grade") + 
  ylab("Verbal Ability") + ylim(0,80) +
  scale_x_continuous(breaks=seq(1,6,by=1)) +
  facet_wrap( ~ id)

#APA style
#ggplot version .. see also http://ggplot.yhathq.com/docs/index.html
ggplot(data = wisclong[wisclong$id<21,], aes(x = grade, y = verb, group = id)) +
  geom_point() + 
  geom_line() +
  xlab("Grade") + 
  ylab("WISC Verbal Score") + ylim(0,100) +
  scale_x_continuous(breaks=seq(1,6,by=1)) +
  #title
  ggtitle("Intraindividual Change in Verbal Ability") +
  #theme with white background
  theme_classic() + 
  #increase font size of axis and point labels
  theme(axis.title = element_text(size = rel(1.5)),
        axis.text = element_text(size = rel(1.2)),
        legend.position = "none")
```
# Saving the plot file. 
See also … http://www.cookbook-r.com/Graphs/Output_to_a_file/

```{r savefile}
#ggsave(filename = default_name(plot), plot = last_plot(), device = default_device(filename), 
#       path = NULL, scale = 1, width = par("din")[1], height = par("din")[2], 
#       units = c("in", "cm", "mm"), dpi = 300, limitsize = TRUE, ...)
ggsave(filename = "wiscverbal.png", width = 5, height = 5, dpi=300)

```




